Interisland movements of woodland caribou in Newfoundland

### 2020-02-19 paper/manuscript.Rmd
reflow P1 few edits
reflow

## Abstract

## Introduction
Caribou are exceptional swimmers. Ample evidence exists that caribou swim in streams, rivers, and lakes during migration (Leblond et al. 2016) to avoid predators (Bergerud 1985) and access islands during calving (Bergerud et al. 1990). Despite a large number of coastal and island caribou herds, limited evidence exists suggesting that caribou swim in the ocean as readily as they swim in freshwater (see Table 1). A key exception is Peary caribou swimming up to 2.6km between the Queen Elizabeth Islands (Miller 1995; Miller 2002). For caribou living on small islands in the ocean, the same notions for why caribou swim in freshwater may also explain why caribou swim in the ocean (Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and predator avoidance (Jeffery et al. 2007) have been proposed as potential explanations.

Swimming for quadrupedal terrestrial mammals can be thought of as a set of trade-offs. Costs include high energy expenditure and risk of drowning. Specifically, even though caribou are adept swimmers, the energetic expenditure associated with swimming for quadrupedal mammals is significantly higher than walking or running (Fish 1993). Additionally, drowning is also possible (Miller & Gunn 1985) although the proportion of caribou that die from drowning is unknown. Swimming between islands or from the mainland to nearby islands can be beneficial. Among the most important benefits are predator avoidance and seeking under-foraged areas.

Woodland caribou are well known to disperse to remote locations, including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987). Moving to remote locations is a viable anti-predator strategy for caribou because it reduces detection and encounter rates by predators. For instance, the use of shoreline habitat in summer was deemed an effective anti-predator calving strategy because caribou avoided predators by using islands and peninsulas or swimming as a flight response (Bergerud et al. 1990). The use of water, and by association occasionally swimming, is therefore part of the fine-scale interactions between caribou and their predators and the use of islands appears to be an effective anti-predator strategy.
	Here, we describe visual observation of a single swimming event by a caribou in the North Atlantic Ocean. This observation led us to investigate the prevalence of swimming of GPS radio-collared caribou in the Fogo Island archipelago in Newfoundland, Canada. We posit two potential mechanisms explaining the trade-offs associated with swimming in the North Atlantic Ocean: forage limitation and the associated density-dependent habitat selection and predator avoidance.

@@

## Discussion
We observed one swimming event in the field as well as 86 remotely sensed swimming events from three individual caribou over a three-year period. In general, caribou swam more frequently to larger islands. Fogo, Perry, and Eastern Indian islands are the largest, and the most commonly occupied. While caribou often swim across streams, rivers, and lakes (Leblond et al. 2016), few observations exist of caribou swimming in the ocean (see Table 1). Of the existing or inferred observations of caribou swimming in the ocean, none appear to be as prolific as the number and frequency of swimming events we observed. We surmise that for caribou living on small islands, the primary drivers of swimming are likely related to forage limitation and predator avoidance.

>>>>

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
2016) to avoid predators (Bergerud 1985) and access islands during calving
(Bergerud et al. 1990). Despite a large number of coastal and island caribou
herds, limited evidence exists suggesting that caribou swim in the ocean as
readily as they swim in freshwater (see Table 1). A key exception is Peary
caribou swimming up to 2.6 km between the Queen Elizabeth Islands (Miller 1995;
Miller 2002). Swimming for quadrupedal terrestrial mammals can be thought of as
a set of trade-offs. Costs include high energy expenditure and risk of drowning.
Specifically, even though caribou are adept swimmers, the energetic expenditure
associated with swimming for quadrupedal mammals is significantly higher than
walking or running (Fish 1993). Additionally, drowning is also possible (Miller
& Gunn 1985) although the proportion of caribou that die from drowning is
unknown. Swimming between islands or from the mainland to nearby islands can be
beneficial. Among the most important benefits are predator avoidance and seeking
under-foraged areas.

We describe visual observation of a single swimming event by a caribou in the
North Atlantic Ocean. This observation led us to investigate the prevalence of
swimming of GPS radio-collared caribou in the Fogo Island archipelago in
Newfoundland, Canada. We posit two potential mechanisms explaining the
trade-offs associated with swimming in the North Atlantic Ocean: forage
limitation and the associated density-dependent habitat selection and predator
avoidance. For caribou living on small islands in the ocean, the same notions
for why caribou swim in freshwater may also explain why caribou swim in the
ocean (Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and
predator avoidance (Jeffery et al. 2007) have been proposed as potential
explanations.

@@

## Discussion

Woodland caribou are well known to disperse to remote locations, including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987). Moving to remote locations is a viable anti-predator strategy for caribou because it reduces detection and encounter rates by predators. For instance, the use of shoreline habitat in summer was deemed an effective anti-predator calving strategy because caribou avoided predators by using islands and peninsulas or swimming as a flight response (Bergerud et al. 1990). The use of water, and by association occasionally swimming, is therefore part of the fine-scale interactions between caribou and their predators and the use of islands appears to be an effective anti-predator strategy.

We observed one swimming event in the field as well as 86 remotely sensed swimming events from three individual caribou over a three-year period. In general, caribou swam more frequently to larger islands. Fogo, Perry, and Eastern Indian islands are the largest, and the most commonly occupied. While caribou often swim across streams, rivers, and lakes (Leblond et al. 2016), few observations exist of caribou swimming in the ocean (see Table 1). Of the existing or inferred observations of caribou swimming in the ocean, none appear to be as prolific as the number and frequency of swimming events we observed. We surmise that for caribou living on small islands, the primary drivers of swimming are likely related to forage limitation and predator avoidance.
